---
title: "Assignment3"
author: "RF"
date: "3/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
p_load(ggplot2, rethinking)
```

```{r}
getwd()
locpath=getwd()
setwd(locpath)
```
## Multivariate linear models

In this assignment you will investigate the relation between different aspects of IQ and symptom severity in ASD (as measured by ADOS). The assignment is meant to make you practice on linear models, multiple predictors and interactions.

The data you will use is the outcome of the psychological testing of the children you analyzed in methods 3.
Data: https://www.dropbox.com/s/hom2qnmp9hyfwmu/Assignment3Data.csv?dl=0
The data consists of ChildID, gender, age, diagnosis (ASD), symptom severity (ADOS), several aspects of IQ (NonVerbal/Perceptual,Verbal,Social) and the visit at which the tests were administered. The questions should be answered by focusing on visit 1.

The questions you will have to answer are:

1. Assess the relation between symptom severity and IQ (all three IQs separately)(focus on visit 1 and children with ASD) and report the model (how did you build it, which priors, data story), a plot of the model, a couple of lines describing the quality of the model and interpreting the results (discussed in lecture). P.S. Should you scale? P.P.S. Don't forget to motivate your priors. P.P.P.S. At least one plot for results and a plot for quality of each model (here and in the next questions) would be appreciated.
1.1. Verbal IQ and ADOS
1.2. Non Verbal IQ and ADOS
1.3. Social IQ and ADOS

centralizing: removes the correlation between alfas and betas because the alfas are centralized around zero. 
scaling: non-linear transformation? (scale function in R does both scaling and centralizing)
scaling function:
scale(x, center = TRUE, scale = TRUE) (x = object/matrix, if center = TRUE, the centering is done by subtracting the column means(omitting NA's) of x from their corresponding columns. If scale = TRUE the scaling is done by dividing the (centered) columns of x by their standard deviations)
a change of 1 sd in IQ will mean 1 sd of change in verbalIQ
NIQ: outcome 
VIQ: predictor 


QUESTION 1: Assessing the relation between symptom severity (ADOS) and IQ (verbal, non-verbal and social)

*Verbal IQ and ADOS*
```{r}
#plot to check if they covary
plot(VerbalIQ ~ ADOS, data = d2)


#Providing starting values for MAP
start = list(
  mu = mean(d2$VerbalIQ),
  sigma = sd(d2$VerbalIQ)
)

#checking the max value of VerbalIQ
max(d2$VerbalIQ)

#fitting model
m1 = map(
  alist(
    VerbalIQ ~ dnorm(mu, sigma),
    mu <- a + b*ADOS,
    a ~ dnorm(18, 10), #based on the start values. prior of 18 because it is the mean of VerbalIQ. SD of verbalIQ is 7.6, and we leave a bit more space for it to deviate by putting 10 as SD here.
    b ~ dnorm(0, 5), #conservative prior of 0 (we don't know what to expect), but with SD of 5 - this is more agnostic
    sigma ~ dunif(0, 33) #measure of error. determined from the range (0-33) of VerbalIQ
    ),
data = d2)

#table of estimates (variance-covariance matrix)
precis(m1, corr = TRUE)

#superimpose MAP values for mean vIQ over the actual data:
plot(VerbalIQ ~ ADOS, data = d2)
abline(a = coef(m1) ["a"], b = coef(m1) ["b"])

#Extracting some samples from the model to show uncertainty
  post <- extract.samples(m1)
  post[1:5,] #inspecting first 5 rows
  
  #Adding the lines
  #extracting all 34 cases in d2
  N <- 34
  dN <- d2[1:N, ]
  mN <- map(
    alist(
      VerbalIQ ~ dnorm(mu, sigma),
      mu <- a+b*ADOS,
      a ~ dnorm(18, 10),
      b ~ dnorm(0, 5),
      sigma ~ dunif(0, 33)
    ), data=dN)

#Plotting 20 of these lines
  #extract 20 samples from the posterior
  post <- extract.samples(mN, n=20)

  #display raw data and sample size
  plot(dN$ADOS, dN$VerbalIQ,
       xlim=range(d2$ADOS), ylim=range(d2$VerbalIQ),
       col=rangi2, xlab="ADOS", ylab="Verbal IQ")
  mtext(concat("N = ", N))
  
  #plot the lines, with transparency
  for(i in 1:20)
    abline(a=post$a[i], b=post$b[i], col=col.alpha("black", 0.3)) #loops over all 20 lines, using abline to display each

#____________________________________________________#
#Contour instead of lines
  
  mu <- link(m1) #link takes the map model fit, samples from the posterior distribution, and computes mean for each case in the data abd sample from the posterior distribution.
  str(mu)
  #We want a distribution of mean for each unique weight value on the horizontal axis:
      #define sequence of weights to compute predictions for. these values will be on the horizontal axis.
      ADOS.seq <- seq(from=-5, to=25, by=1)
      
      #use link to compute mu for each sample in the posterior and for each weight in the weight.seq
      mu <- link(m1, data=data.frame(ADOS=ADOS.seq))
      str(mu)

    #Summarising the distribution for each weight value
    #summarise the distribution of mu
    mu.mean <- apply(mu, 2, mean) #compute the mean of each column (dimension "2") of the matrix mu.
    mu.HPDI <- apply(mu, 2, HPDI, prob=0.89)
    
#--Prediction intervals--
  #Generating a 89% prediction interval for actual heights, not just the average height (mean); incorporating SD and its uncertainty.
  sim.ADOS <- sim(m1, data=list(ADOS=ADOS.seq))
  str(sim.ADOS) #Matrix that contains simulated heights, not distributions of plausible average height as earlier.
  
  #Summarising them by using apply
  ADOS.PI <- apply(sim.ADOS, 2, PI, prob=0.89)


    #plot raw data
    plot(VerbalIQ ~ ADOS, d2, col=col.alpha(rangi2, 0.5))
    
    #draw MAP line
    lines(ADOS.seq, mu.mean)
    
    #draw HPDI region for line
    shade(mu.HPDI, ADOS.seq)
    
    #draw PI region for simulated heights
    shade(ADOS.PI, ADOS.seq)
```
We built a model using MAP and alist, with Verbal IQ as the outcome with a normal distribution, and ADOS as the predictor with a normal distribution.

We have chosen not to scale (standardize) our data so that we are able to interpret the data on a natural scale.

Motivation for priors:
For alpha we chose a prior of 18 because it is the mean of VerbalIQ. SD of verbalIQ is 7.6, and we leave a bit more space for it to deviate by putting 10 as SD here.
For beta we chose a conservative prior of 0 because we don't know what to expect, but with a SD of 5 as this is more agnostic.
For sigma, which has a uniform distribution, we chose the range 0-33, as that is the range of possible scores of verbal IQ

From the variance-covriance matrix we can tell that for every increase in ADOS score, vIQ drops with 1.1, and that 89% of the posterior probability lies between -1.4 and -0.8, which suggests that beta values well below -1.4 and above -0.8 are highly incompatible with these data and this model. Furthermore, we notice from the table that an ASD child with an ADOS score of 0 has a vIQ of 33, and that 95% of plausible vIQs lie within 10 vIQ scores (2 sigmas) of the mean vIQ.
We also see that the two parameters alpha and beta are almost negatively correlated, meaning they carry almost the same information. We assess that this is not a problem in this simple model (in more complex models we can use centering to adress the problem).

We interpret all this as there being a negative relationship between symptom severity and verbal IQ in ASD children, meaning that children with higher scores in ADOS show lower scores in verbal IQ. It is not surprising to see a relationship between verbal IQ and symptom severity, as children with e.g. severe autism often show a lack of language/communication skills. It should also be considered that IF (we did not research this) a high score of ADOS is based on (among other things) poor language skills, the relationship is already given from this.

We produced four plots: One of the data, one of the data and the model, one of the data and the model with uncertainty, and lastly one of the data with shaded regions of uncertainty. 
The plots show a negative relationship between ADOS and verbal IQ.
From the third plot, it can be seen that there is greater uncertainty at extreme measures of ADOS, e.g. ADOS = 0, which is probably because of the very small portion of participants in this area (e.g. child 74 that has an ADOS score of 0). From the last plot we visualize the data and model as follows: The line is the MAP estimate of the mean vIQ at each ADOS value. The narrow shaded interval is the distribution of mu (a + b*ADOS). The wider shaded region represents the region within which the model expects to find 89% of actual vIQs in the group of ASD children at each ADOS value. The shaded regions are very wide, which shows that there is a lot of uncertainty, consequently affecting the quality of the model. This is probably due to low sample size (only 34 participants in visit 1 with ASD).
__________________________


*Non-verbal IQ and ADOS*
```{r}
#plot to check if they covary
plot(NonVerbalIQ ~ ADOS, data = d2)

#Providing starting values for MAP
start = list(
  mu = mean(d2$NonVerbalIQ),
  sigma = sd(d2$NonVerbalIQ)
)

#checking the max value of NonVerbalIQ
max(d2$NonVerbalIQ)
# mean of NonVerbalIQ
mean(d2$NonVerbalIQ)
    # 27
# sd of NonVerbalIQ
sd(d2$NonVerbalIQ)
    # 5.4

m2 = map(
  alist(
    NonVerbalIQ ~ dnorm(mu, sigma),
    mu <- a + b*ADOS,
    a ~ dnorm(27, 10), #based on the start values. prior of 27 because it is the mean of NonVerbalIQ. SD of NonverbalIQ is 5.4, and we leave a bit more space for it to deviate by putting 10 as SD here. (also because it is closer to the 2 times the std.dev.??)
    b ~ dnorm(0, 5), #conservative prior of 0 (we don't know what to expect), but with SD of 5 - this is more agnostic
    sigma ~ dunif(0, 42) #measure of error. determined from the range (0-42) of NonVerbalIQ
    ),
data = d2)

#table of estimates (variance-covariance matrix)
precis(m2, corr = TRUE)

#superimpose MAP values for mean vIQ over the actual data:
plot(NonVerbalIQ ~ ADOS, data = d2)
abline(a = coef(m2) ["a"], b = coef(m2) ["b"])

#Extracting some samples from the model to show uncertainty
 
  #Adding the lines
  #extracting all 34 cases from d2
  N <- 34
  dN <- d2[1:N, ]
  mN <- map(
    alist(
      NonVerbalIQ ~ dnorm(mu, sigma),
      mu <- a+b*ADOS,
      a ~ dnorm(27, 10),
      b ~ dnorm(0, 5),
      sigma ~ dunif(0, 42)
    ), data=dN)
#Plotting 20 of these lines
  #extract 20 samples from the posterior
post <- extract.samples(mN, n=20)

  #display raw data and sample size
plot(dN$ADOS, dN$NonVerbalIQ,
       xlim=range(d2$ADOS), ylim=range(d2$NonVerbalIQ),
       col=rangi2, xlab="ADOS", ylab="Non Verbal IQ")
  mtext(concat("N = ", N))
  
  #plot the lines, with transparency
for(i in 1:20)
  abline(a=post$a[i], b=post$b[i], col=col.alpha("black", 0.4)) #loops over all 20 lines, using abline to display each

#____________________________________________________#
#Contour instead of lines
  
  mu <- link(m2) #link takes the map model fit, samples from the posterior distribution, and computes mean for each case in the data abd sample from the posterior distribution.
  str(mu)
  #We want a distribution of mean for each unique weight value on the horizontal axis:
      #define sequence of weights to compute predictions for. these values will be on the horizontal axis.
      ADOS.seq <- seq(from=-5, to=25, by=1)
      
      #use link to compute mu for each sample in the posterior and for each weight in the weight.seq
      mu <- link(m2, data=data.frame(ADOS=ADOS.seq))
      str(mu)

    #Summarising the distribution for each weight value
    #summarise the distribution of mu
    mu.mean <- apply(mu, 2, mean) #compute the mean of each column (dimension "2") of the matrix mu.
    mu.HPDI <- apply(mu, 2, HPDI, prob=0.89)
    
#--Prediction intervals--
  #Generating a 89% prediction interval for actual heights, not just the average height (mean); incorporating SD and its uncertainty.
  sim.ADOS <- sim(m2, data=list(ADOS=ADOS.seq))
  str(sim.ADOS) #Matrix that contains simulated heights, not distributions of plausible average height as earlier.
  
  #Summarising them by using apply
  ADOS.PI <- apply(sim.ADOS, 2, PI, prob=0.89)


    #plot raw data
    plot(NonVerbalIQ ~ ADOS, d2, col=col.alpha(rangi2, 0.5))
    
    #draw MAP line
    lines(ADOS.seq, mu.mean)
    
    #draw HPDI region for line
    shade(mu.HPDI, ADOS.seq)
    
    #draw PI region for simulated heights
    shade(ADOS.PI, ADOS.seq)
```
By using MAP and alist we built a model with NonVerbalIQ as the outcome with a normal distribution, and ADOS as the predictor with a normal distribution.

As in the previous model, We have chosen not to scale (standardize) our data so that we are able to interpret the data on a natural scale.

Motivation for priors: 
For alpha we chose a prior of 27 because it is the mean of NonVerbalIQ. SD of NonVerbalIQ is 5.4, and we leave a bit more space for it to deviate by putting 10 as SD here.(also because it is closer to the 2 times the std.dev.??)
For beta we chose a conservative prior of 0 because we don't know what to expect, but with a SD of 5 as this is more agnostic.
For sigma, which has a uniform distribution, we chose the range 0-42, as that is the range of possible scores of non verbal IQ

From the variance-covariance matrix of estimates we can see that for every increase in ADOS score, NonVerbalIQ drops with 0.6. We can also tell that 89% of the posterior probability lies between -0.9 and -0.4, which suggests that beta values well below -0.9 and above -0.4 are highly incompatible with these data and this model. Furthermore, we notice from the table that an ASD child with an ADOS score of 0 has a NonVerbalIQ of 36 (intercept), and that 95% of plausible NonVerbalIQs lie within 8 NonVerbalIQ scores (2 sigmas) of the mean NonVerbalIQ.
We can also see that the two parameters alpha and beta are almost negatively correlated, meaning they carry almost the same information. We assess that this is not a problem in this simple model (in more complex models we can use centering to adress the problem).

We interpret this as there being a slightly negative relationship between symptom severity and non verbal IQ in ASD children, meaning that children with higher scores in ADOS show lower scores in non verbal IQ. Children with ASD often show poor language skills, but this need not necessarily affect their non-verbal intelligence. This could be the reason for this slightly lesser relationship between symptom severity and non-verbal IQ.

We produced four plots: One of the data, one of the data and the model, one of the data and the model with uncertainty, and lastly one of the data with shaded regions of uncertainty, respectively. From the third plot, we now see a little less uncertainty at extreme values, as there is not as wide a spread in the samples.
From the last plot we visualize the data and model as we did for verbal IQ. The shaded regions are a little less wide then with verbal iq as the predictor, but there is still some uncertainty.
________________________


*Social IQ and ADOS*
```{r}
#plot to check if they covary
plot(SocialIQ ~ ADOS, data = d2)

#Providing starting values for MAP
start = list(
  mu = mean(d2$SocialIQ),
  sigma = sd(d2$SocialIQ)
)

#checking the max value of SocialIQ
max(d2$SocialIQ)
    # 105
# mean of SocialIQ
mean(d2$SocialIQ)
    # 78
# sd of SocialIQ
sd(d2$SocialIQ)
    # 11.3

m3 = map(
  alist(
    SocialIQ ~ dnorm(mu, sigma),
    mu <- a + b*ADOS,
    a ~ dnorm(78, 15), #based on the start values. prior of 78 because it is the mean of SocialIQ. SD of SocialIQ is 11.3, and we leave a bit more space for it to deviate by putting 15 as SD here.
    b ~ dnorm(0, 5), #conservative prior of 0 (we don't know what to expect), but with SD of 5 - this is more agnostic
    sigma ~ dunif(0, 105) #measure of error. determined from the range (0-105) of SocialIQ
    ),
data = d2)

#table of estimates
precis(m3, corr = TRUE)

#superimpose MAP values for mean vIQ over the actual data:
plot(SocialIQ ~ ADOS, data = d2)
abline(a = coef(m3) ["a"], b = coef(m3) ["b"])

#Extracting some samples from the model to show uncertainty
  

  #Adding the lines
  #extracting all 34 cases in d2
  N <- 34
  dN <- d2[1:N, ]
  mN <- map(
    alist(
      SocialIQ ~ dnorm(mu, sigma),
      mu <- a+b*ADOS,
      a ~ dnorm(78, 15),
      b ~ dnorm(0, 5),
      sigma ~ dunif(0, 105)
    ), data=dN)
#Plotting 20 of these lines
  #extract 20 samples from the posterior
post <- extract.samples(mN, n=20)

  #display raw data and sample size
plot(dN$ADOS, dN$SocialIQ,
       xlim=range(d2$ADOS), ylim=range(d2$SocialIQ),
       col=rangi2, xlab="ADOS", ylab="Social IQ")
  mtext(concat("N = ", N))
  
  #plot the lines, with transparency
for(i in 1:20)
  abline(a=post$a[i], b=post$b[i], col=col.alpha("black", 0.4)) #loops over all 20 lines, using abline to display each

#____________________________________________________#
#Contour instead of lines
  
  mu <- link(m3) #link takes the map model fit, samples from the posterior distribution, and computes mean for each case in the data abd sample from the posterior distribution.
  str(mu)
  #We want a distribution of mean for each unique weight value on the horizontal axis:
      #define sequence of weights to compute predictions for. these values will be on the horizontal axis.
      ADOS.seq <- seq(from=-5, to=25, by=1)
      
      #use link to compute mu for each sample in the posterior and for each weight in the weight.seq
      mu <- link(m3, data=data.frame(ADOS=ADOS.seq))
      str(mu)

    #Summarising the distribution for each weight value
    #summarise the distribution of mu
    mu.mean <- apply(mu, 2, mean) #compute the mean of each column (dimension "2") of the matrix mu.
    mu.HPDI <- apply(mu, 2, HPDI, prob=0.89)
    
#--Prediction intervals--
  #Generating a 89% prediction interval for actual heights, not just the average height (mean); incorporating SD and its uncertainty.
  sim.ADOS <- sim(m3, data=list(ADOS=ADOS.seq))
  str(sim.ADOS) #Matrix that contains simulated heights, not distributions of plausible average height as earlier.
  
  #Summarising them by using apply
  ADOS.PI <- apply(sim.ADOS, 2, PI, prob=0.89)


    #plot raw data
    plot(SocialIQ ~ ADOS, d2, col=col.alpha(rangi2, 0.5))
    
    #draw MAP line
    lines(ADOS.seq, mu.mean)
    
    #draw HPDI region for line
    shade(mu.HPDI, ADOS.seq)
    
    #draw PI region for simulated heights
    shade(ADOS.PI, ADOS.seq)
```
Again we used MAP and alist to build a model with SocialIQ as the outcome with a normal distribution, and ADOS as the predictor with a normal distribution.

Again, we are not scaling.

Motivation for priors: 
For alpha we chose a prior of 78 because it is the mean of SocialIQ. SD of SocialIQ is 11.3, and we leave a bit more space for it to deviate by putting 15 as SD here.
For beta we chose a conservative prior of 0 because we don't know what to expect, but with a SD of 5 as this is more agnostic.
For sigma, which has a uniform distribution, we chose the range 0-105, as that is the range of possible scores of non verbal IQ

From the variance-covariance matrix we can tell that for every increase in ADOS score, SocialIQ drops with 1.5. We can also tell that 89% of the posterior probability lies between -1.99 and -1.02, which suggests that beta values well below -2 and above -1 are highly incompatible with these data and this model. Furthermore, we notice from the table that an ASD child with an ADOS score of 0 has a SocialIQ of 99 (intercept), and that 95% of plausible SocialIQs lie within 17 SocialIQ scores (2 sigmas) of the mean SocialIQ.

We can also see that the two parameters alpha and beta are almost negatively correlated, meaning they carry almost the same information. We assess that this is not a problem in this simple model (in more complex models we can use centering to adress the problem).

We interpret this as there being a negative relationship between symptom severity and social IQ in ASD children, meaning that children with higher scores in ADOS show lower scores in social IQ. 

We produced four plots: One of the data, one of the data and the model, one of the data and the model with uncertainty, and lastly one of the data with shaded regions of uncertainty. 
The plots show a negative relationship between ADOS and verbal IQ.
From the third plot, it can be seen that there is greater uncertainty at extreme measures of ADOS. The shaded regions in the fourth plot are very wide, which shows that there is a lot of uncertainty, consequently affecting the quality of the model.

2. Do the different aspects of IQ account for different portions of the variance in ADOS? 
```{r}

# exploration by precis plot (variance plot?)
model <- map(
    alist(
        ADOS ~ dnorm( mu , sigma ) ,
        mu <- a + bN*NonVerbalIQ + bV*VerbalIQ + bS*SocialIQ ,
        a ~ dnorm( 14 , 5 ) ,
        bN ~ dnorm(27, 10) ,
        bV ~ dnorm(18, 10) ,
        bS ~ dnorm(78, 15) ,
        sigma ~ dunif( 0 , 21 )
),
    data = d2 )
precis( model )
precis_plot(precis(model))

#pairs plot
pairs( ~ NonVerbalIQ + VerbalIQ + SocialIQ ,
    data=d2 , col=rangi2 )
# the pairs plot show the variance shared by variables (not entirely sure)
# along the diagonal, the variables are labeled. In each scatterplot off the diagonal, the vartical axis variable is the variable labeled in the same column e.g. the two scatterplots in the first row are NonVerbalIQ (vertical) against VerbalIQ (horizontal), and then NonVerbalIQ (vertical) against SocialIQ (horizontal)
# In this plot - if the scatter form a pretty diagonal straight line, it is because the two variables are so correlated (negatively or positively) to such a degree that they are nearly redundant - this means that either of them helps predicting the last variable, but neither helps much once you already know the other. 
# we can also compute the correlation betweent them with cor()


```


2.1. Does it make sense to have all IQ measures in the same model? First write a few lines answering the question and motivating your answer, including a discussion as to what happens when you put all of them in the same model. Then build a model following your answer. If your answer is "no", you are not free, you still have to answer: are there alternative ways of answering the question?

There is doubt about whether the different measures of IQ are even different, because some argue that they are very correlated. If we choose to put very correlated variables in a model together, we face the problem of multicollinearity, meaning that the variables will most likely cancel each other out. If we are not aware of this possible multicollinearity before we run the model - when we see that the posterior distribution says that a very large range of posterior variables are plausible - we could mistakenly read it as saying that neither predictor is important, even if all the variables are in reality strongly associated with the outcome variable. This is because multiple regression controls for all other variables, and so what we are asking is whether adding another predictor will tell us more than if we only use one predictor.
If we believe that the different measures of IQ are very correlated, it will not make sense to put them all in a model as predictors.
We will first test for multicollinearity, and from that figure out what predictors to put in our model.

```{r}
#Checking correlation by doing a pairs plot
pairs( ~ ADOS + NonVerbalIQ + VerbalIQ + SocialIQ ,
    data=d2 , col=rangi2 )
# the pairs plot show the variance shared by variables
# along the diagonal, the variables are labeled. In each scatterplot off the diagonal, the vartical axis variable is the variable labeled in the same column e.g. the two scatterplots in the first row are NonVerbalIQ (vertical) against VerbalIQ (horizontal), and then NonVerbalIQ (vertical) against SocialIQ (horizontal)
# In this plot - if the scatter form a pretty diagonal straight line, it is because the two variables are correlated (negatively or positively) to such a degree that they are nearly redundant - this means that either of them helps predicting the last variable, but neither helps much once you already know the other.

```
We have done a pairs plot to see the correlation between the outcome variable and the variables, as well as the correlation between the variables inbetween. 
It seems from the plot that there is a negative relation between the outcome variable ADOS and each of the variables, nV, V and S, but having already familiarized ourselves with the data, we expected this negative relationship. However, looking at the variables it seems that there is a tendency to correlation between each of the variables, although we are careful not to assume too much as there are so many outliers in the data. 
In order to check the actual correlation, we compute the correlation between each of the variables with the cor() function.


```{r}

# correlation between VerbalIQ and NonVerbalIQ
cor(d2$VerbalIQ , d2$NonVerbalIQ)
# correlation between VerbalIQ and SocialIQ
cor(d2$VerbalIQ , d2$SocialIQ)
# correlation between NonVerbalIQ and SocialIQ
cor(d2$NonVerbalIQ , d2$SocialIQ)
# how the variables correlate 
```

v~nV = 0.67
V~S = 0.6
nV~S = 0.37

The correlation between Verbal and nonVerbal IQ is 0.67, and the correlation between Verbal and Social IQ is 0.6. The correlation between nonVerbal and Social IQ is 0.37. Because the correlation between nonVerbal and Social IQ is the lowest, we choose these two variables as our predictors in our model. 
If we were interested in making further analysis of the correlation we could compute a precis plots of the different combinations of variables to see whether the beta estimates move and by how much, and how much their standard deviation changes.

2.2. Build the model, assess its quality, write a few lines interpreting the results.

First we computed a model with the unstandardized values as we have done previously. However, the precis plot is hard to read and we are not quite sure how to interpret it. Therefore, we have made the model with the standardized data.
```{r}
# model with nV and S (UNSTANDARDIZED)
model_nV_S_1 <- map(
    alist(
        ADOS ~ dnorm( mu , sigma ) ,
        mu <- a + bN*NonVerbalIQ + bS*SocialIQ ,
        a ~ dnorm(14 , 5) , 
        bN ~ dnorm(27, 10) , 
        bS ~ dnorm(78, 15) ,
        sigma ~ dunif( 0 , 21 ) #the 89% uncertainty. HAS to be positive
),
    data = d2 )
precis( model_nV_S_1)
precis_plot(precis(model_nV_S_1))

d = read.csv("Assignment3Data.csv")

#Creating data frame with only data from visit 1 for children with ASD
d2 = subset(d, ASD == 1 & Visit == 1)

d2 = d2[complete.cases(d2),] #Making new data frame without NAs

# standardizing the data (centering and scaling)
d2_s = scale(d2, center = TRUE, scale = TRUE)
d2_s = as.data.frame(d2_s)

# model with nV and S (STANDARDIZED)
model_nV_S <- map(
    alist(
        ADOS ~ dnorm( mu , sigma ) ,
        mu <- a + bN*NonVerbalIQ + bS*SocialIQ ,
        a ~ dnorm( -1.07e-16 , 1 ) , #we put mean at ~0, which we can see on the precis plot, where a is right at 0 (because we have standardized the data)
        bN ~ dnorm(2.66e-16, 1) , 
        bS ~ dnorm(-1.36e-16, 1) ,
        sigma ~ dunif( 0 , 1.54 ) #the 89% uncertainty. HAS to be positive
),
    data = d2_s )
precis( model_nV_S )
precis_plot(precis(model_nV_S))

```
What's the value of knowing each predictor after already knowing all the other predictors? 
There seems to be some value in knowing each predictor, as they do not overlap with zero. Although the two predictors seem to be sharing some variance as shown by the overlap. 


3. Let's now include also the TD children. Does it make sense to ask whether IQ and ADOS are related? Motivate your answer. In any case, if you wanted to build a model to answer that question, which model would you build? Run the model, assess its quality, write a few lines interpreting the results.



```{r}
# data from visit 1 with all TD & ASD children
d1 = subset(d, Visit == 1)
# Making new data frame without NAs
d1 = d1[complete.cases(d1),]
# standardizing the data (centering and scaling)
d1_s = scale(d1, center = TRUE, scale = TRUE)
d1_s = as.data.frame(d1_s)

#make model with data from both
model_nV_S1 <- map(
    alist(
        ADOS ~ dnorm( mu , sigma ) ,
        mu <- a + bN*NonVerbalIQ + bS*SocialIQ ,
        a ~ dnorm( -1.07e-16 , 1 ) , #we put mean at ~0, which we can see on the precis plot, where a is right at 0 (because we have standardized the data)
        bN ~ dnorm(2.66e-16, 1) , 
        bS ~ dnorm(-1.36e-16, 1) ,
        sigma ~ dunif( 0 , 1.54 ) #the 89% uncertainty. HAS to be positive
),
    data = d1_s )

# call link without specifying new data
# so it uses original data
mu <- link( model_nV_S1 )
# summarize samples across cases
mu.mean <- apply( mu , 2 , mean )
mu.PI <- apply( mu , 2 , PI )
# simulate observations
# again no new data, so uses original data
model.sim <- sim( model_nV_S , n=1e4 )
model.PI <- apply( model.sim , 2 , PI )

# plot predictions against observed 
plot( mu.mean ~ d1_s$ADOS , col=rangi2 , ylim=range(mu.PI) ,
    xlab="Observed ADOS" , ylab="Predicted ADOS" )
abline( a=0 , b=1 , lty=2 )
for ( i in 1:nrow(d1_s) )
    lines( rep(d1_s$ADOS[i],2) , c(mu.PI[1,i],mu.PI[2,i]) ,
        col=rangi2 )
```
From the posterior predictive plot we can see that the model over-predicts the children with low ADOS scores (mostly TD children), while under-predicting children with higher ADOS scores (ASD children). It can also be seen that there seems to be two populations, one with low ADOS scores and one with high ADOS scores, which makes sense because there are two populations: children with ASD and TD children. Therefore, it does not make sense to ask whether IQ and ADOS are related in this model when TD children are included.The reason why it doesn't make sense in this model is that the model doesn't account for the two populations. To account for this, we could proceed to make a model that includes Diagnosis as a beta so the model knows to treat these differently.

Attempt to make model with Diagnosis as beta:
```{r}
# model with ADOS as an additional beta 
# 
# model_nV_S1 <- map(
#     alist(
#         ADOS ~ dnorm( mu , sigma ) ,
#         mu <- a + bN*NonVerbalIQ + bS*SocialIQ + bD *Diagnosis ,
#         a ~ dnorm( -1.07e-16 , 1 ) , #we put mean at ~0, which we can see on the precis plot, where a is right at 0 (because we have standardized the data)
#         bN ~ dnorm(2.66e-16, 1) , 
#         bS ~ dnorm(-1.36e-16, 1) ,
#         bD ~ dbinom() #??
#         sigma ~ dunif( 0 , 1.54 ) #the 89% uncertainty. HAS to be positive
# ),
#     data = d1_s )

```

4. Let's discuss contents:
4.1. You have three scores for IQ, do they show shared variance? Is that the same in TD and ASD? What does that tell us about IQ?
4.2. You have explored the relation between IQ and ADOS. How do you explain that relation from a cognitive perspective? N.B. You can present alternative hypotheses.
```{r}
######
# dataframe with only TD children
d_TD = subset(d, ASD == 0 & Visit == 1)
d_TD = d_TD[complete.cases(d_TD),] #Making new data frame without NAs

# correlation between VerbalIQ and NonVerbalIQ
cor(d_TD$VerbalIQ , d_TD$NonVerbalIQ)
# correlation between VerbalIQ and SocialIQ
cor(d_TD$VerbalIQ , d_TD$SocialIQ)
# correlation between NonVerbalIQ and SocialIQ
cor(d_TD$NonVerbalIQ , d_TD$SocialIQ)
```
The three scores for IQ are somewhat correlated, but it is not the same in TD and ASD children. There is a higher correlation between the IQs in ASD children and a lower correlation between the IQs in the TD children. For both TD and ASD children, verbal and nonverbal IQ are the most correlated (0.67 for ASD and 0.55 for TD). However, social IQ is not very correlated with the other two types of IQ in TD children, and in ASD children, social IQ and non-verbal IQ are also not very correlated. We understand this as a suggestion that the general intelligence hypothesis can't be supported, as the different types of IQ are not all correlated in neither ASD nor TD children.

We suspect that we observe a weaker correlation between non-verbal/verbal IQ and social IQ in ASD children due to the relationship between autism and Theory of Mind - it has been found that autists generally have a harder time with Theory of Mind, which relates to social IQ. 



note: nonverbal + verbal IQ has been shown to be positively correlated with grey matter volume. 



__________________________________________________

notes to self from class:
quality check:
 - the plot with the lines that show mean uncerternty is the best to show the quality, but not when we move onward from bivariate regressions to multivariate regressions
 - predictive posterior plor: the model "curve" and the raw data "curve" tell us something about what the model does and does not capture. In the ados model we can see that the data has a higher peak than the model, but because the model has a Gaussian expectation of the data we are actually restraining the model - if it should have a higher peak the model would fit worse on the "sides" so the model show the best compromise. 
 
if the lines overlap with 0, the variable does not give any new information. or do they? 
we can also start by looking at how the individual variables correlate to one another pairwise.

v~nV = 67
V~S = 0.6
nV~S = 0.37
kenneth: look at the individual beta values for individual variables once we know the others and see how much it moves. 
some evidence of a g factor, but yet the model seems to describe different phenomena. There is ilttle shared variance between nV and S, meaning that the variance between v~nV and the variance between V~S is not the same variance - and therefore maybe not the g factor... 
correlations:
how are we going to quantify how much the variables correlate with each other? 
- of the model (ADOS~VIQ+NIQ+SIQ) (same as: y ~ a + Bn*NIQ + Bv*VIQ + Bs*SIQ)
- maybe look at the precis(m, corr = TRUE)
- does the correlation matrix tell how the variables are correlated 
- how toquantify how much the variables correlate with each other? Make a map model and oen variable will be y, another x???? 
- Normally we only look at the correlation of varibales
- if the model gived problems / irregularities, we look at the correlation of beta values?

- when does it make sense to add a second predictor?
what does it mean if they are all far away from xero/the mean. 


__________________


_______________________________________________________________________________________________________

#Notes to self:
height ~ weight
b = 0.9
1 up in weight, the beta value increases -> 1 extra kilo means 0.9 cm taller

vIQ ~ ADOS
b = -1.1
1 up in ADOS, the beta value decreases -> 1 extra value of ADOS means 1.1 less vIQ = the worse the symptom severity, the lower vIQ

89% of the posterior probability lies between -1.4 and -0.8 = suggests that beta values well below -1.4 and above -0.8 are highly incompatible with these data and this model

mean of alpha indicates that a person of ADOS 0 should have a vIQ of 33

mean of sigma indicates the width of the distribution of vIQ around the mean. 95% of plausible vIQs lie within 10 vIQ scores (2 sigmas) of the mean vIQ.

from the variance-covariance matrix we see that the two parameters alpha and beta are almost negatively correlated -> they carry almost the same information. not a problem in this simple mode (in more complex models we can use centering to adress the problem).

plot:
multiple regression lines sampled from the posterior. it can be seen that there is greater uncertainty at extreme measures of ADOS, e.g. ADOS = 0


W

Variable of interest: Verbal IQ and whether it is influences by ADOS. Verbal IQ has a Gaussian distribution.

based on the start values. prior of 18 because it is the mean of VerbalIQ. SD of verbalIQ is 7.6, and we leave a bit more space for it to deviate by putting 10 as SD here.
    b ~ dnorm(0, 5), #conservative prior of 0 (we don't know what to expect), but with SD of 5 - this is more agnostic
    sigma ~ dunif(0, 33) #measure of error. determined from the range (0-33) of VerbalIQ
    
about the plot with predicted and observed aDOS:
##kigget på predictions (observed and predicted ADOS overfor hinanden). gør forskel fordi mere diverse scores når alle er inkluderet. alle TD children har ADOS = 0, hvilket ikke giver mening at inkludere - gør modellen dårligere. hvis de ikke har en ADOS score, så kan man jo ikke sige om IQ and ADOS er related. negatively affect predictive power fordi TD children har lav ADOS score - ender med at over og underpredicte for de to kategorier af børn. to forskellige populations.   hvis bruge diagnosis som beta bliver modellen bedre fordi den kan tage højde for det. 
